Sparse spectral methods for partial differential equations on spherical caps

### 2021-04-16 papers/spherical-caps/revised/response-to-referees.tex

\centerline{\textbf{Response to Referee 2}}

\begin{itemize}[parsep=1em,leftmargin=1em]

\item \comment{One shortcoming of this paper is that it has no discussion of the condition number of the linear systems to be solved. I suspect that the condition numbers can get large, but if so, can the matrices be preconditioned in a similar way the Olver?Townsend ultraspherical spectral method (so that it has bounded condition number)?}

Added small subsection and plots to show the condition numbers of each m-block of Laplacian operator, along with a preconditioned Laplacian (where the preconditioner $P$ is the inverse of the diagonal of the Laplacian). While the Laplacian does not have bounded condition number, the trivial preconditioning keeps it bounded.

### 2021-04-16 papers/spherical-caps/revised/spherical-caps.tex
Finished making the changes to paper, with new figs for condition numbers.

\begin{figure}[t]
	\centering % <-- added
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[scale=0.3]{sparsity-of-laplacian}
		\centering
		\caption{The Laplace-Beltrami operator $\Delta^{(1)}_W$}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[scale=0.3]{sparsity-of-rho2laplacian}
		\centering
		\caption{The $\rho^2$-factored Laplace-Beltrami operator $D_\varphi^{(0)} \: W_\varphi^{(1)} + T^{(0)\to(1)} \: T_W^{(1)\to(0)} \: (D_\theta)^2$}
	\end{subfigure}\hfil % <-- added

	\medskip
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[scale=0.3]{sparsity-of-helmholtz}
		\centering
		\caption{A variable coefficient Helmholtz operator}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[scale=0.3]{sparsity-of-biharmonic}
		\centering
		\caption{The Biharmonic operator $\calB_W^{(2)}$} 
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of (differential) operator matrices, showing their sparsity. For (c), the weighted variable coefficient Helmholtz operator is $\Delta^{(1)}_W + k^2 \: T^{(0)\to(1)} \: V({J_x^{(0)}}^\top, {J_y^{(0)}}^\top, {J_z^{(0)}}^\top) \: T_W^{(1)\to(0)}$ for $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 200$.}
	\label{fig:sparsity}
\end{figure}

@@

@@

\begin{figure}[t]
	\begin{subfigure}{0.5\textwidth}
	\includegraphics[scale=0.25]{poisson-u=wyexpx-N=60}
	\centering
	%\label{fig:solution-poisson}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.3]{solutionblocknorms-poisson-epsilonfun-N=200}
        	%\label{fig:solutionblocknorms-poisson}
	\end{subfigure}
	\caption{Left: The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y,z) = - 2 e^x y z (2+x) + \genjacw^{(1,0)}(z) e^x (y^3 + z^2 y - 4xy - 2y)$. Right: The norms of each block of the computed solution of the Poisson equation with right hand side function $f(x,y,z) = \norm{\xvec - (\epsilon + 1 / \sqrt{3}) \; (1,1,1)^\top }$ for different $\epsilon$ values. This indicates spectral convergence.}
	\centering
	\label{fig:poisson}
\end{figure}

\begin{figure}[t]
	\begin{subfigure}{0.5\textwidth}
	\includegraphics[scale=0.25]{helmholtz-f=wyexpx-N=60-n=300}
	\centering
	%\label{fig:solution-poisson}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.3]{solutionblocknorms-helmholtz-varyingk-N=200}
        	%\label{fig:solutionblocknorms-poisson}
	\end{subfigure}
	\caption{Left: The computed solution to $\Delta u + k^2 \: v \: u = f$ with zero boundary conditions with $f(x,y,z) = y e^x (z - \alpha) $, $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 100$. Right: The norms of each block of the computed solution of the Helmholtz equation with the right hand side function $f(x,y,z) = 1$ and the same function $v(x,y,z)$, for various $k$ values. This indicates spectral convergence.}
	\centering
	\label{fig:helmholtz}
\end{figure}

\begin{figure}[t]
	\centering % <-- added
	\includegraphics[scale=0.35]{complexity-Nend=1000}
	\caption{Time in seconds to build and solve the system $\big[\DeltaS + v(x,y,z) \big] \: u(x,y,z) = f(x,y,z)$, for a rotationally invariant $v(x,y,z) = v(z)$. This demonstrates that the approach is roughly of order $\bigO(N^2)$, where $N$ is the degree to which we approximate the solution (the number of unknowns is then $(N+1)^2$). Here, we used $f = -2e^{x}yz(2+x) + (z - \alpha) e^{x} (y^3 + z^2 y - 4xy - 2y)$ and $v(x,y,z) = v(z) = \cos(z)$.}
	\centering
	\label{fig:complexity}
\end{figure}

\begin{figure}[t]
	\begin{subfigure}{0.5\textwidth}
	\includegraphics[scale=0.15]{biharmonic-f=erf-N=80-n=50}
	\centering
	%\label{fig:solution-biharmonic}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
	\includegraphics[scale=0.3]{solutionblocknorms-biharmonic-expfun-N=200}
	\centering
        	%\label{fig:solutionblocknorms-biharmonic}
	\end{subfigure}
	\caption{Left: The computed solution to $\Delta^2 u = f$ with zero Dirichlet and Neumann boundary conditions with $f(x,y,z) = (1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))) \rho(z)^2$. Right: The norms of each block of the computed solution of the biharmonic equation with the right hand side function $f(x,y,z) = \exp(-\epsilon((x-x_0)^2 + (y-y_0)^2 + (z-z_0)^2))$ where  $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$, for various $\epsilon$ values. This demonstrates algebraic convergence.}
	\centering
	\label{fig:biharmonic}
\end{figure}
% END FIGURES

@@

@@

>>>>

\begin{figure}[tp]
	\centering % <-- added
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.5]{sparsity-of-laplacian}
		\centering
		\caption{The Laplace-Beltrami operator $\Delta^{(1)}_W$}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.5]{sparsity-of-rho2laplacian}
		\centering
		\caption{The $\rho^2$-factored Laplace-Beltrami operator $D_\varphi^{(0)} \: W_\varphi^{(1)} + T^{(0)\to(1)} \: T_W^{(1)\to(0)} \: (D_\theta)^2$}
	\end{subfigure}\hfil % <-- added

	\medskip
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.5]{sparsity-of-helmholtz}
		\centering
		\caption{A variable coefficient Helmholtz operator}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.5]{sparsity-of-biharmonic}
		\centering
		\caption{The Biharmonic operator $\calB_W^{(2)}$} 
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of (differential) operator matrices, showing their sparsity. For (c), the weighted variable coefficient Helmholtz operator is $\Delta^{(1)}_W + k^2 \: T^{(0)\to(1)} \: V({J_x^{(0)}}^\top, {J_y^{(0)}}^\top, {J_z^{(0)}}^\top) \: T_W^{(1)\to(0)}$ for $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 200$.}
	\label{fig:sparsity}
\end{figure}

@@

\subsection{Stability of the Laplacian operator}

\begin{figure}[tp]
	\centering
	\begin{subfigure}[b]{0.55\textwidth}
		\includegraphics[scale=0.4]{condition-numbers-laplacian-N=200}
	\end{subfigure}
	\hfill%

	\begin{subfigure}[b]{0.55\textwidth}
		\includegraphics[scale=0.4]{condition-numbers-max-laplacian}
	\end{subfigure}
	\hfill%
	\caption{Plots of the condition number for the Laplacian operator matrix $\Delta^{(1)}_W$ and the preconditioned matrix $P^{-1} \; \Delta^{(1)}_W$ where $P$ is the matrix of the diagonal of $\Delta^{(1)}_W$. Left: The condition numbers of each diagonal block of $\Delta^{(1)}_W$ for $N = 200$, as well as those for $P^{-1} \; \Delta^{(1)}_W$. Right: The maximum condition number of all diagonal blocks of the Laplacian and the preconditioned Laplacian operators, as the degree $N$ increases.}
	\label{fig:conditionnumbers}
\end{figure}

For a given problem $A\vec{x} = \vec{b}$, the condition number is a measure of how accurate solving the system will be (ignoring any numerical rounding errors), and is given as the condition number of the matrix $A$: 
\begin{align*}
	\kappa(A) := \norm{A}_2 \norm{A^{-1}}_2.
\end{align*}
Recall that the Laplacian operator is block-diagonal. One can therefore look at the condition numbers of each block on the diagonal as proxy for the condition number of the Laplacian (and hence the Poisson equation), where graphical representations are seen in \ref{fig:conditionnumbers} for the $\Delta^{(1)}_W$ operator matrix. As the degree $N$ increases, the condition numbers of these blocks becomes increasingly large (at least, the condition number for the first block where $m=0$ does, which will be the largest for each $N$). Fortunately however, by applying a trivial preconditioning matrix similar to that seen for the ultraspherical method \cite{olver2013fast}, we can significantly bound the condition numbers -- this is also seen in \ref{fig:conditionnumbers}. The preconditioning matrix chosen is $P^{-1}$, where $P$ is the matrix of the diagonal of $\Delta^{(1)}_W$.

@@

\begin{figure}[tp]
	\centering
	\begin{subfigure}{0.85\textwidth}
		\includegraphics[scale=0.4]{poisson-u=wyexpx-N=60}
	\end{subfigure}
	\hfill%

	\begin{subfigure}{0.55\textwidth}
		\includegraphics[scale=0.4]{solutionblocknorms-poisson-epsilonfun-N=200}
	\end{subfigure}
	\hfill%
	\caption{Top: The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y,z) = - 2 e^x y z (2+x) + \genjacw^{(1,0)}(z) e^x (y^3 + z^2 y - 4xy - 2y)$. Bottom: The norms of each block of the computed solution of the Poisson equation with right hand side function $f(x,y,z) = \norm{\xvec - (\epsilon + 1 / \sqrt{3}) \; (1,1,1)^\top }$ for different $\epsilon$ values. This indicates spectral convergence.}
	\label{fig:poisson}
\end{figure}

% END FIGURES

@@

\begin{figure}[tp]
	\centering
	\begin{subfigure}{0.85\textwidth}
		\includegraphics[scale=0.4]{helmholtz-f=wyexpx-N=60-n=300}
	\end{subfigure}

	\begin{subfigure}{0.55\textwidth}
		\includegraphics[scale=0.4]{solutionblocknorms-helmholtz-varyingk-N=200}
	\end{subfigure}
	\caption{Top: The computed solution to $\Delta u + k^2 \: v \: u = f$ with zero boundary conditions with $f(x,y,z) = y e^x (z - \alpha) $, $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 100$. Bottom: The norms of each block of the computed solution of the Helmholtz equation with the right hand side function $f(x,y,z) = 1$ and the same function $v(x,y,z)$, for various $k$ values. This indicates spectral convergence.}
	\label{fig:helmholtz}
\end{figure}

\begin{figure}[tp]
	\centering % <-- added
	\includegraphics[scale=0.4]{complexity-Nend=1000}
	\caption{Time in seconds to build and solve the system $\big[\DeltaS + v(x,y,z) \big] \: u(x,y,z) = f(x,y,z)$, for a rotationally invariant $v(x,y,z) = v(z)$. This demonstrates that the approach is roughly of order $\bigO(N^2)$, where $N$ is the degree to which we approximate the solution (the number of unknowns is then $(N+1)^2$). Here, we used $f = -2e^{x}yz(2+x) + (z - \alpha) e^{x} (y^3 + z^2 y - 4xy - 2y)$ and $v(x,y,z) = v(z) = \cos(z)$.}
	\label{fig:complexity}
\end{figure}

@@

\begin{figure}[tp]
	\centering
	\begin{subfigure}{0.85\textwidth}
		\includegraphics[scale=0.4]{biharmonic-f=erf-N=80-n=50}
	\end{subfigure}
	\hfill%

	\begin{subfigure}{0.55\textwidth}
		\includegraphics[scale=0.4]{solutionblocknorms-biharmonic-expfun-N=200}
	\end{subfigure}
	\hfill%
	\caption{Left: The computed solution to $\Delta^2 u = f$ with zero Dirichlet and Neumann boundary conditions with $f(x,y,z) = (1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))) \rho(z)^2$. Right: The norms of each block of the computed solution of the biharmonic equation with the right hand side function $f(x,y,z) = \exp(-\epsilon((x-x_0)^2 + (y-y_0)^2 + (z-z_0)^2))$ where  $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$, for various $\epsilon$ values. This demonstrates algebraic convergence.}
	\label{fig:biharmonic}
\end{figure}
